---
title: "2026 Summer  BIO 2100 Midterm Creation and Answers"
subtitle: "33 points"
bibliography: ref_2026_2100_summer_midterm.bib
format:
  html:
    embed-resources: true
    toc: true
    toc-float: true
    toc-depth: 6
    code-fold: false
editor:
  visual:
    wrap: 72
  source:
    wrap: 72
---

## Instructions! Read this first.

The exam is open note, open course website, open class-related code
repositories (mine and those you produced in class). However, you may
not get external help ( from other students, AI, directed internet
searches, etc.). Please update the statement below to acknowledge these
instructions (and that you can use git).

I, INSERT YOUR NAME HERE, promise to not seek external help on the exam.
I understand any academic issues will result in a D or F on the exam or
in the class and be reported to the Dean of Students. Infractions will
also result in me being unable to obtain a letter of recommendation from
the department for professional school applications.

The exam should be committed and pushed to github by Thursday,
6/25/2026, at 10:15 AM. You can also commit and push along the way, but
please mark your final commit (when you are finished) with the comment
**Finished - ready to grade!.**

For all questions, the statement is based on what you might hear from a
colleague (e.g., "How do I analyze this?"). This is like when you get a
call to review a paper. Please answer questions succinctly but with
necessary details on all required steps/concerns you might need to
consider.

While the exam is based on a realistic scenario (with citations), some
data and scenarios are fictional and contrived for the exam (i.e., don't
quote all this as fact!).

Good luck!

## Background

"The use of cryptic coloration for visual camouflage (eg background
matching) is widespread in the animal kingdom, occurring when the
coloration and patterning of an organism resembles the surrounding
substrate" [@messas2025]. Multiple traits, such as coloration, the
ability to perceive, compare, and choose habitat based on color, and the
ability to change color (within or across life stages), may all
contribute to background matching and be under selective pressure.

These issues were considered in a series of studies focused on the
garden cross spider (*Araneus diadematus*). This species can range from
light yellow to dark grey in general coloration. They may engage in
crypsis to avoid predators or to allow them to ambush their prey. These
spiders build orb webs, then retreat and hide to wait and ambush
captured prey.

[![Figure 1 from Messas et al 2025. Camouflage and color variation of
the European garden spider Araneus diadematus. A) The resting position
of a spider upon a leaf, evidencing a better match to brown than green
backgrounds. B-C) Examples of how the coloration of spiders and their
respective host leaves vary from dark to pale brown. Photo credits: Yuri
Fanchini
Messas.](Picture1.jpg){fig-alt="Picture of spiders camouflaged on leaves."}](https://academic.oup.com/view-large/figure/524067907/araf057_fig1.jpg)

## 1

Spiders (n=30) were presented with a choice experiment where half the
space was covered with green paper and half with brown paper. Each side
also contained a leaf of corresponding color. Spiders were released in
the center of the space and position was recorded after 24 hours.

24 spiders chose to build their web on the green side.

How would you use statistics to analyze the data. Make sure you include

-   null hypothesis (1 point)
  - H~0~: Spiders have no preference for color (i.e., p=.5)

-   alternative hypothesis (1 point)
  - H~A~: Spiders have a preference for color (i.e., p$\neq$.5)

-   explanation for test you will use (1.5 point)
  - I have two outcomes that I am comparing to a set proportion of success, so
  I will use a binomal test.

-   results from statistical test (1 point)

```{r}
binom.test(24,30)
```


-   clear explanation of how results relate to your stated hypotheses
    (1.5 point)
      - since p=.0014 (<.05), I reject my null hypothesis. There is 
      evidence for preference of color.

## 2

A colleague thinks your analysis is wrong. They instead want to focus on
the number choosing the brown side because that is what was observed
most commonly in the field, even though brown leaves are rare. Do you
need to re-do the analysis? Why or why not? Explain.

These results are from a controlled experiment with a 50/50 split, so brown 
leaves being rare in the field does not impact them.  Also, for binomial data
I can choose either outcome as the success or failure - it won't impact the 
results.

## 3

In the same experiment, 14 spiders chose to hide under the green leaf.

How would you use statistics to analyze the data. Make sure you include

-   null hypothesis (1 point)
  - H~0~: Spiders have no preference for leaf color (i.e., p=.5)

-   alternative hypothesis (1 point)
  - H~A~: Spiders have a preference for leaf color (i.e., p$\neq$.5)


-   explanation for test you will use (1.5 point)
  - I have two outcomes that I am comparing to a set proportion of success, so
  I will use a binomal test.

-   results from statistical test (1 point)

```{r}
binom.test(14,30)
```

-   clear explanation of how results relate to your stated hypotheses
    (1.5 point)
      - since p=.855 (>.05), I fail to reject my null hypothesis. There is no 
      evidence for preference of leaf color.
    

## 4

Scientists also considered the ability of spiders to change their color
to match their leaves. The dataset below contains information on color
change (a unitless measure; **change** column) for spiders kept in brown
habitats for 14 days. Is there any evidence spiders can change colors?

```{r}
color_change_brown <- read.csv("color_change_brown.csv", stringsAsFactors = T)
```

Analyze the data. Make sure you include

-   null hypothesis (1 point)
  - H~0~: There is no change in spider color ($\mu_{change}=0$)

-   alternative hypothesis (1 point)
  - H~A~: There is change in spider color ($\mu_{change}\neq0$)

-   explanation for test you will use (1.5 point)
  - I will use a t-test as I am comparing continuous data to a set value and I 
  don't know the variance of the underlying population. Color should also be
  a hidden average of multiple factors (genetics, diet, development, etc)

-   results from statistical test (1 point)

    ```{r}
    t.test(color_change_brown$change)
    ```

-   clear explanation of how results relate to your stated hypotheses
    (1.5 point)
      - Given t~14~=-1.673 (p=0.12), I fail to reject the null hypothesis. There
      is no evidence of color change.

## 5

A colleague is confused. They note the average color change was -1.38
units and many of the spiders did change color. Why do you need to test
for color change if those facts are true?

Spiders may change color for a number of reasons. However, what we observed may
be due to sampling error or measurement issues. There was no consistent evidence 
of change.

## 6

A larger study included color change in spiders kept on green and brown
substrates (the **substrate** column. Data are at

```{r}
color_change_all <- read.csv("color_change_all.csv", stringsAsFactors = T)
```

Construct an appropriate plot to showcase how color change differed
among substrates.

Many options here.

```{r}
library(ggplot2)
ggplot(color_change_all, aes(x=substrate, y=change, fill=substrate))+
  geom_boxplot(alpha=.4, outlier.shape=NA, width=.5)+
  geom_jitter(aes(color=substrate), width = .08, size = 3, alpha=0.8)+
  labs(
    x="Substrate",
    y= "Color change (unitless)",
    title="Spider color change by substrate type"
  ) + 
  theme_minimal(base_size=13)+
  theme(legend.position="none")
```



```{r}
ggplot(color_change_all, aes(x=substrate, y=change, fill=substrate))+
  geom_jitter(aes(color=substrate), width = .08, size = 3, alpha=0.8)+
  labs(
    x="Substrate",
    y= "Color change (unitless)",
    title="Spider color change by substrate type"
  ) + 
  theme_minimal(base_size=13)+
  theme(legend.position="none")
```

```{r}
library(Rmisc)
color_change_summary <- summarySE(color_change_all, measurevar = "change", groupvars = "substrate", na.rm=T)
ggplot(color_change_summary, aes(x=substrate, y=change, fill=substrate))+
  geom_col()+
  geom_errorbar(aes(ymin=change-ci, ymax=change+ci), size=1.5)
```

```{r}
ggplot(color_change_summary, aes(x=substrate, y=change, fill=substrate))+
  geom_point()+
  geom_errorbar(aes(ymin=change-ci, ymax=change+ci), size=1.5)
```


