---
title: "2026 Summer  BIO 2100 Midterm Creation and Answers"
subtitle: "33 points"
bibliography: ref_2026_2100_summer_midterm.bib
format:
  html:
    embed-resources: true
    toc: true
    toc-float: true
    toc-depth: 6
    code-fold: false
editor:
  visual:
    wrap: 72
  source:
    wrap: 72
---

## Instructions! Read this first.

The exam is open note, open course website, open class-related code
repositories (mine and those you produced in class). However, you may
not get external help ( from other students, AI, directed internet
searches, etc.). Please update the statement below to acknowledge these
instructions (and that you can use git).

I, INSERT YOUR NAME HERE, promise to not seek external help on the exam.
I understand any academic issues will result in a D or F on the exam or
in the class and be reported to the Dean of Students. Infractions will
also result in me being unable to obtain a letter of recommendation from
the department for professional school applications.

The exam should be committed and pushed to github by Thursday,
6/25/2026, at 10:15 AM. You can also commit and push along the way, but
please mark your final commit (when you are finished) with the comment
**Finished - ready to grade!.**

For all questions, the statement is based on what you might hear from a
colleague (e.g., "How do I analyze this?"). This is like when you get a
call to review a paper. Please answer questions succinctly but with
necessary details on all required steps/concerns you might need to
consider.

While the exam is based on a realistic scenario (with citations), some
data and scenarios are fictional and contrived for the exam (i.e., don't
quote all this as fact!).

Good luck!

## Background

"The use of cryptic coloration for visual camouflage (eg background
matching) is widespread in the animal kingdom, occurring when the
coloration and patterning of an organism resembles the surrounding
substrate" [@messas2025]. Multiple traits, such as coloration, the
ability to perceive, compare, and choose habitat based on color, and the
ability to change color (within or across life stages), may all
contribute to background matching and be under selective pressure.

These issues were considered in a series of studies focused on the
garden cross spider (*Araneus diadematus*). This species can range from
light yellow to dark grey in general coloration. They may engage in
crypsis to avoid predators or to allow them to ambush their prey. These
spiders build orb webs, then retreat and hide to wait and ambush
captured prey.

[![Figure 1 from Messas et al 2025. Camouflage and color variation of
the European garden spider Araneus diadematus. A) The resting position
of a spider upon a leaf, evidencing a better match to brown than green
backgrounds. B-C) Examples of how the coloration of spiders and their
respective host leaves vary from dark to pale brown. Photo credits: Yuri
Fanchini
Messas.](Picture1.jpg){fig-alt="Picture of spiders camouflaged on leaves."}](https://academic.oup.com/view-large/figure/524067907/araf057_fig1.jpg)

## 1

Spiders (n=30) were presented with a choice experiment where half the
space was covered with green paper and half with brown paper. Each side
also contained a leaf of corresponding color. Spiders were released in
the center of the space and position was recorded after 24 hours.

24 spiders chose to build their web on the green side.

How would you use statistics to analyze the data. Make sure you include

-   null hypothesis (1 point)

-   alternative hypothesis (1 point)

-   explanation for test you will use (1.5 point)

-   results from statistical test (1 point)

-   clear explanation of how results relate to your stated hypotheses
    (1.5 point)

## 2

A colleague thinks your analysis is wrong. They instead want to focus on
the number choosing the brown side because that is what was observed
most commonly in the field, even though brown leaves are rare. Do you
need to re-do the analysis? Why or why not? Explain.

## 3

In the same experiment, 14 spiders chose to hide under the green leaf.

How would you use statistics to analyze the data. Make sure you include

-   null hypothesis (1 point)

-   alternative hypothesis (1 point)

-   explanation for test you will use (1.5 point)

-   results from statistical test (1 point)

-   clear explanation of how results relate to your stated hypotheses
    (1.5 point)

## 4

Scientists also considered the ability of spiders to change their color
to match their leaves.

Had to read in original data and make a change column for ease

```{r}
color_change <- read.csv("colour_change_rnl_data.csv", stringsAsFactors = T)

```

```{r}
library(reshape2)
color_change_wide <- dcast(color_change, spider_id + substrate~day)
```

```{r}
color_change_wide$change <- color_change_wide$`14`-color_change_wide$`0`
```

only use brown leaves here

```{r}
color_change_brown <- color_change_wide[color_change_wide$substrate=="brown",
                                        c("spider_id", "substrate", "change")]
color_change_all <- color_change_wide[,
                                        c("spider_id", "substrate", "change")]
```

```{r}
write.csv(color_change_brown, "color_change_brown.csv", row.names = F)
write.csv(color_change_all, "color_change_all.csv", row.names = F)
```

The dataset below contains information on color change (a unitless
measure; **change** column) for spiders kept in brown habitats for 14
days. Is there any evidence spiders can change colors?

```{r}
color_change_brown <- read.csv("color_change_brown.csv", stringsAsFactors = T)
```

Analyze the data. Make sure you include

-   null hypothesis (1 point)

-   alternative hypothesis (1 point)

-   explanation for test you will use (1.5 point)

-   results from statistical test (1 point)

    ```{r}
    t.test(color_change_brown$change)
    ```

-   clear explanation of how results relate to your stated hypotheses
    (1.5 point)

## 5

A colleague is confused. They note the average color change was -1.38
units and many of the spiders did change color. Why do you need to test
for color change if those facts are true?

## 6

A larger study included color change in spiders kept on green and brown
substrates (the **substrate** column. Data are at

```{r}
color_change_all <- read.csv("color_change_all.csv", stringsAsFactors = T)
```

Construct an appropriate plot to showcase how color change differed
among substrates.
