---
title: "2026 Spring BIO 2100 Final Creation and Answers"
subtitle: "33 points"
bibliography: ref_2026_2100_spring_final.bib
format:
  html:
    embed-resources: true
    toc: true
    toc-float: true
    toc-depth: 6
    code-fold: false
editor:
  visual:
    wrap: 72
  source:
    wrap: 72
---

## Instructions! Read this first.

The exam is open note, open course website, open class-related code
repositories (mine and those you produced in class). However, you may
not get external help ( from other students, AI, directed internet
searches, etc.). Please update the statement below to acknowledge these
instructions (and that you can use git).

I, INSERT YOUR NAME HERE, promise to not seek external help on the exam.
I understand any academic issues will result in a D or F on the exam or
in the class and be reported to the Dean of Students. Infractions will
also result in me being unable to obtain a letter of recommendation from
the department for professional school applications.

The exam should be committed and pushed to github by Monday, 5/18/2026,
at 10:00 AM. You can also commit and push along the way, but please mark
your final commit (when you are finished) with the comment **Finished -
ready to grade!.**

For all questions, the statement is based on what you might hear from a
colleague (e.g., "How do I analyze this?"). This is like when you get a
call to review a paper. Please answer questions succinctly but with
necessary details on all required steps/concerns you might need to
consider.

While the exam is based on a realistic scenario (with citations), some
data and scenarios are fictional and contrived for the exam (i.e., don't
quote all this as fact!).

Good luck!

## Background

Community science is a growing area where people who are not
traditionally considered scientists assist with research projects. This
can include providing data, assisting with project implementation, and
analyzing data [@gosnell2023]. Community scientists can expand data
collection efforts beyond what a single lab can accomplish and greatly
enrich projects.

A recent example focused on using social media posts to determine the
distribution of ornate spotted rays (*Aetomylaeus vespertilio*)
[@araujo2020].

[![Photo of a rarely seen Ornate eagle ray (Aetomylaeus vespertilio),
taken at Landaa Giraavaru, Baa Atoll, Maldives (February 2018).
SunnyMart, CC BY-SA 4.0
\<https://creativecommons.org/licenses/by-sa/4.0\>, via Wikimedia
Commons](Ornate_eagle_ray_(Aetomylaeus_vespertilio)_Baa_Atoll,_Maldives_(2018).jpg)](https://commons.wikimedia.org/wiki/File:Ornate_eagle_ray_(Aetomylaeus_vespertilio)_Baa_Atoll,_Maldives_(2018).jpg)

## 1

Previous research indicated eagle rays are nocturnal feeders, with 80%
of their feeding activity occurring at night. The collected images
included 76 shots of rays feeding, with 50 of those shot in daylight
hours. Does this match expectations?

Make sure you include (6 points)

-   null hypothesis (1 point)
-   alternative hypothesis (1 point)
-   explanation for test you will use (1.5 point)
-   results from statistical test (1 point)
-   clear explanation of how results relate to your stated hypotheses
    (1.5 point)

## 2

Scientists used imaging tools to gather extra information from the the
images. Note different images provided different pieces of data due to
the perspective. For example, they were able to use some pictures to
determine both sex (male, female, unknown) and wing width for each
individual. Data is available at

```{r}
set.seed(29)
ray_data <- data.frame(
  sex = c(rep("male", 42), 
           rep("female", 28), 
           rep("unknown", 63)), 
  width_cm = c(round(rnorm(42, 165, 5), 2), 
               round(rnorm(28,218, 5),2), 
               round(rnorm(63,47,5),2)))
write.csv(ray_data, "ray_data.csv", row.names = F)
rm(ray_data)
```

```{r}
ray_data <- read.csv("ray_data.csv", stringsAsFactors = T)
```

Use the data to determine if wing width differs among sexes. Make sure
you include (6 points)

-   null hypothesis (1 point)

-   alternative hypothesis (1 point)

-   explanation for test you will use (1.5 point)

-   results from statistical test (1 point)

-   clear explanation of how results relate to your stated hypotheses
    (1.5 point)

## 3

Plot the data from question 2.


## 4

The dataset also gave a chance to consider how social the rays are.
Scientists considered images of each sex to determine if they were
photographed alone or in groups.

|                | Male | Female | Unknown |
|----------------|------|--------|---------|
| **Alone**      | 38   | 6      | 14      |
| **In a group** | 4    | 22     | 49      |

Use the data to determine if sociality differs among sexes. Make sure
you include (6 points)

-   null hypothesis (1 point)

-   alternative hypothesis (1 point)

-   explanation for test you will use (1.5 point)

-   results from statistical test (1 point)

-   clear explanation of how results relate to your stated hypotheses
    (1.5 point)


## 5

Ornate eagle rays have very long tails - they can be up to 2-3 m long. A
few images were found that allowed both the tail and the wing width to
be measured. Is there any relationship among these body parts? Data is
available at

```{r}
set.seed(29)
wing_tail_data <- data.frame(
    wing_width_cm = c(round(rnorm(52, 175, 25),2)))
wing_tail_data$tail_length_cm <- wing_tail_data$wing_width_cm*2.4+runif(52)*10
write.csv(wing_tail_data, "wing_tail_data.csv", row.names = F)
rm(wing_tail_data)
```

```{r}
wing_tail_data <- read.csv("wing_tail_data.csv", stringsAsFactors = T)
```

Make sure you include (6 points)

-   null hypothesis (1 point)

-   alternative hypothesis (1 point)

-   explanation for test you will use (1.5 point)

-   results from statistical test (1 point)

-   clear explanation of how results relate to your stated hypotheses
    (1.5 point)


## 6

In general, do you see any issues with using this data or the analysis
you completed? Please note at least potential concerns or issues with
any of the above datasets or analysis.
