---
title: "2026 Spring BIO 2100 Midterm"
subtitle: "33 points"
bibliography: ref_2026_2100_spring_midterm.bib
format:
  html:
    embed-resources: true
    toc: true
    toc-float: true
    toc-depth: 6
    code-fold: false
editor:
  visual:
    wrap: 72
  source:
    wrap: 72
---

## Instructions! Read this first.

The exam is open note, open course website, open class-related code
repositories (mine and those you produced in class). However, you may
not get external help ( from other students, AI, directed internet
searches, etc.). Please update the statement below to acknowledge these
instructions (and that you can use git).

I, INSERT YOUR NAME HERE, promise to not seek external help on the exam.
I understand any academic issues will result in a D or F on the exam or
in the class and be reported to the Dean of Students. Infractions will
also result in me being unable to obtain a letter of recommendation from
the department for professional school applications.

The exam should be committed and pushed to github by Monday, 3/28/2026,
at 9:55 AM. You can also commit and push along the way, but please mark
your final commit (when you are finished) with the comment **Finished -
ready to grade!.**

For all questions, the statement is based on what you might hear from a
colleague (e.g., "How do I analyze this?"). This is like when you get a
call to review a paper. Please answer questions succinctly but with
necessary details on all required steps/concerns you might need to
consider.

While the exam is based on a realistic scenario (with citations), some
data and scenarios are fictional and contrived for the exam (i.e., don't
quote all this as fact!).

Good luck!

## Background

Where do bees go in the winter? And how do they survive?

[![Bombus impatiens, the common eastern bumblebee. Photographed in New
Jersey. By Mousebelt - Own work, CC BY-SA 4.0,
https://commons.wikimedia.org/w/index.php?curid=98700493](BumbleBee.jpg){fig-alt="Picture of bee (Bombus impatiens) on flower"}](https://commons.wikimedia.org/w/index.php?curid=98700493)

New common eastern bumblee (*Bombus impatiens*) queens (gynes) nest
underground after mating in the fall. They start new colonies the
following spring. Given that, how do they survive rain and snow during
their hibernation, which may last for months?

A lab accident led to the discovery that bees may be able to survive
submersion [@hibernat].

## 1

During an experiment focused on the impact of insecticides on common
eastern bumblebees, a graduate student was keeping bee queens in
soil-filled tubes in the refrigerator (to mimic winter conditions). One
day she noticed some of the tubes were filled with water. She quickly
removed the bees but expected them to be dead. Instead, 20 out of 21
were still moving.

This may be an example of where statistics is not needed. Why?

No bees were expected to survive, but almost all (20/21) did; this
suggests a clear pattern that does not requires p-value/test-based
analysis.

## 2

An editor suggested a short comment noting these results should have a
p-value (probably a bad idea..). However, how would you use statistics
to analyze the data. Make sure you include

-   null hypothesis (1 point)

    -   H~O~: all the bees should have died due to water, so
        p\[survival\]=0

-   alternative hypothesis (1 point)

    -   H~A~: p\[survival\] $\neq$ 0

-   explanation for test you will use (1.5 point)

    -   I would use a binomial test because i have data from 2
        categories and am testing a proportion

-   results from statistical test (1 point)

    ```{r}
    binom.test(20,21,0)
    ```

-   clear explanation of how results relate to your stated hypotheses
    (1.5 point)

    -   Since the p-value is less than .05 (p = \<.01), I reject the
        null hypothesis.\
        It appears that bees are able to survive underwater.

## 3

A fellow student is confused about your analysis. They focused on
mortality (opposite of survival) and found a signal of `r 1/21`
<!--new trick! render this (or look at the html) and note it gives you the answer! this is an in-line code chunk  -->
in the data and took that as p. Why is your value different? Make sure
you explain what a p-value is!

They found the proportion (sometimes call the p parameter) focused on
mortality (we did survival), not the p-value. A p-value is the
likelihood of finding the signal you did or something as or more extreme
if the null hypothesis is true and we sampled the population multiple
times.

## 4

Note during creation: understand the data

```{r}
bee_data <- read.csv("data_flooding.csv", stringsAsFactors = T)
```

```{r}
summary(bee_data)
unique(bee_data$Bee_ID)
```

143 unique bees

```{r}
library(Rmisc)
summarySE(bee_data, groupvars = c("Bee_ID"), measurevar = "Weight")
```

each bee assessed three times

```{r}
summarySE(bee_data, groupvars = c("Tx"), measurevar = "Weight")
```

17 control, 21 in each duration x submersion treatment

```{r}
summarySE(bee_data, groupvars = c("Tx", "Duration", "Time"), measurevar = "Weight")

```

A follow-up study submersion [@rondeau2024] experiment compared survival
under three different conditions.

-   **control** (no water)

-   **floating** in water

-   maintained **underwater** via a plunger

![Three treatments used in the study. From Rondeau and Raine
2026.](rsbl.2023.0609.f001.png){fig-alt="Three test tubes are shown. A bee is in each, and each has dirt on the bottom. One bee is on the dirt (control), one is floating in water that has been added to the tube, and one is submerged in water by a plunger that has been used to cap the tube."}

Bees were initially placed in soil filled tubes and refrigerated. After
7 days, submersion treatments were applied. For the treatments with
water, bees were maintained for 8 hours, 24 hours, or 7 days. A total of
143 bees (17 control, 21 in each duration x treatment condition) were
used. After the experiment, all bees were returned to new tubes with
soil, returned to the refrigerator, then assessed for survival at one,
four, and eight weeks following the beginning of submersion treatments.

For our exam, we will focus on the data collected at eight weeks
following 24 hour submersion treatments. Below code creates a table if
you wish to use the combined data. Note the following columns are used

-   Submersion

    -   treatment (control; floating, underwater =submerged)

-   Weight

    -   weight in mg (measured before the study)

-   Survival

    -   1= survived; 0 = did not survive

```{r}
bee_data <- read.csv("data_flooding.csv", stringsAsFactors = T)

```

```{r}
bee_data <- read.csv("Data_flooding.csv", stringsAsFactors = T)
library(plyr)
bee_data$Submersion <-revalue(bee_data$Submersion, c("ctl"="control")) 
bee_data <- bee_data[bee_data$Time==8 & bee_data$Duration==24 & bee_data$Weight != "NA" &
                       !is.na(bee_data$Duration)|
                       bee_data$Time==8 & bee_data$Submersion=="control" & bee_data$Weight != "NA" &
                        !is.na(bee_data$Duration),]
bee_table <- table(bee_data[bee_data$Time==8 & bee_data$Duration==24|
                       bee_data$Time==8 & bee_data$Submersion=="control","Submersion"],
      bee_data[bee_data$Time==8 & bee_data$Duration==24|
                       bee_data$Time==8 & bee_data$Submersion=="control","Survival"])
colnames(bee_table) <- c("Survived", "Did not survive")
bee_table
```

Analyze the data. Make sure you include

-   null hypothesis (1 point)

    -   H~O~: survival is independent of treatment *OR* proportion
        surviving is the same for each treatment

-   alternative hypothesis (1 point)

    -   H~A~: survival is not independent of treatment *OR* proportion
        surviving is not the same for each treatment

-   explanation for test you will use (1.5 point)

    -   I will use a $\chi^2$ test since I'm comparing proportions among
        various groups.

-   results from statistical test (1 point)

    ```{r}
    chisq.test(bee_table)
    ```

    Since the $\chi^2$ test returned a warning, I checked expected
    outcomes to make sure they met assumptinos

    ```{r}
    chisq.test(bee_table)$expected
    ```

    Given 50% of the groups are less than 5, I used a Fisher test
    instead

    ```{r}
    fisher.test(bee_table)
    ```

-   clear explanation of how results relate to your stated hypotheses
    (1.5 point)

    -   We fail to reject the null hypothesis (p=.8493).

## 5

Data was also collected on the weights of the bees (*Weight* column in
**bee_data**). Plot the weight data to consider differences among
submersion treatments.

Several options here:

```{r}
library(Rmisc)
library(ggplot2)
bee_summary <- summarySE(bee_data[!is.na(bee_data$Submersion),], measurevar="Weight", groupvars="Submersion")
ggplot(bee_summary, aes(x=Submersion, y=Weight)) +
    geom_point(size=3) +
    geom_errorbar(aes(ymin=Weight - ci, ymax=Weight + ci), width=0) +
    ylab("Mass (mg)") +
    xlab("Treatment") +
    ggtitle("Mass by submersion treatment")
```

```{r}
ggplot(bee_data[!is.na(bee_data$Submersion),], aes(y=Weight, x=Submersion, color=Submersion)) +
  geom_jitter() +
  labs(title="Masses of bees used in various treaments",
       subtitle = "There is no apparent difference in weights",
       y= "Bee weight (mg)",
       x= "Treatment")+
  guides(color=F)
```

```{r}
ggplot(bee_data[!is.na(bee_data$Submersion),], aes(x=Weight, fill=Submersion)) +
  geom_histogram() +
  labs(title="Masses of bees used in various treaments",
       subtitle = "There is no apparent difference in weights",
       y= "Bee weight (mg)",
       x= "Treatment")+
  guides(fill=F)+
  facet_wrap(~Submersion, ncol=1)
```

## 6

Another lab group wants to use this data to consider if bees used in the
study differed from typical queen weights. A typical queen at the stage
used in the study weighs 450 mg.

Investigate the question. Make sure you include

-   null hypothesis (1 point)

    -   H~O~: there is no difference in our bees from typical bees *OR*
        mean weight ($\mu_{weight}$) = 450 mg

-   alternative hypothesis (1 point)

    -   H~O~: there is a difference in our bees from typical bees *OR*
        mean weight ($\mu_{weight}$) $\neq$ 450 mg

-   explanation for test you will use (1.5 point)

    -   I will use a t-test to compare continuous data that I don't know
        the standard deviation for (of the population) to a given value

-   results from statistical test (1 point)

    ```{r}
    t.test(bee_data$Weight, mu=450)
    ```

-   clear explanation of how results relate to your stated hypotheses
    (1.5 point)

    -   We reject the null hypothesis (t~48~=16.971, p\<.001). It
        appears the bees in our study are larger.
