---
title: "2024 Fall 78002 Midterm"
bibliography: references.bib
format:
  html:
    embed-resources: true
    toc: true
    toc-float: true
    toc-depth: 6
    code-fold: false
---

## Instructions! Read this first.

The exam is open note, open course website, open class-related code
repositories (mine and those you produced in class). However, you may
not get external help ( from other students, directed internet searches,
etc.). Please update the statement below to acknowledge these
instructions (and that you can use git).

I, INSERT YOUR NAME HERE, promise to not seek external help on the exam.
I understand any academic issues will result in a D or F on the exam or
in the class and be reported to the Dean of Students. Infractions will
also result in me being unable to obtain a letter of recommendation from
the department for professional school applications.

Good luck!

## Background

A recent paper [@piotto2024] found that micropredation (when animals eat
only a portion of a target prey) by gulls may be important to survival
in southern right whale (*Eubalaena australis*) calves in Argentina.
These predation events cause lesions to develop on whales that may
increase susceptibilty to infection; attempts to prevent them may reduce
maternal energy stores and time spent feeding and resting.

A new graduate student in your program is interested in this topic and
how it may be mediated by food availability, especially in urban
systems. In 2023 over a quarter of the world's North Atlantic right
whale population was observed off Long Island [@riley2024], and while
kelp gulls *Larus dominicanus* are not found in New York, a sister
species (great black-backed gull (*Larus marinus*)) is. Great
black-backed gulls are known to be extremely aggressive predators. They
are also known to forage at dumps [@cavanaugh].

## 1

The student collected data on whales off the coast of Long Island using
photos.  The photos showed birds were present in the area and included a few
direct acts of attack. Images of whales showed that 36 out of 117 of
whales have lesions! In general, whales have
lesions 28% of the time for a variety of reasons. 

Analyze the data. Make sure you include

-   null hypothesis (1 point)

-   alternative hypothesis (1 point)

-   explanation for test you will use (1.5 point)

-   results from statistical test (1 point)

-   clear explanation of how results relate to your stated hypotheses (1.5 point)

## 2

In a follow-up study, your friend gets information on gull attacks on
whales in areas with low (near dumps, which attract birds to land),
medium (in open waters), and high (near fishing vessel routes) gull
density (n=200 per activity type). They then monitor the stress of
whales in these areas (through measurements of glucocorticoid levels
(GC, ng/g^-1^)) obtained by collecting whale feces, following work by @lemos2020)
to see if the whales showed any differences in anxiety in these areas.
The data are provided below.


```{r}
whale_stress <- readRDS("whale_stress_lab.RDS")
```

Analyze the data. Make sure you include

-   null hypothesis (1 point)

-   alternative hypothesis (1 point)

-   explanation for test you will use (1.5 point)

-   results from statistical test (1 point)

-   clear explanation of how results relate to your stated hypotheses (1.5 point)

## 3

Provide (the images) 2 options for plotting the data (don't worry about units if 
not provided).
Briefly discuss the pros and cons of each. (6 pts)

# 4

Another measure of stress, TSH levels, was gathered from the population with
high gull activity.  You can access it using:

```{r}
whale_tsh <- readRDS("whale_tsh.RDS")
```

TSH levels should be low (10 ng/g^-1^).  Analyze the data. Make sure you include

-   null hypothesis (1 point)

-   alternative hypothesis (1 point)

-   explanation for test you will use (1.5 point)

-   results from statistical test (1 point)

-   clear explanation of how results relate to your stated hypotheses (1.5 point)


## 5

Another study uses low and high activity levels and compares calf
survival in these areas.

|                 |                     |                      |
|-----------------|---------------------|----------------------|
|                 | *Low gull activity* | *High gull activity* |
| *Calf survived* | 31                  | 21                   |
| *Calf died*     | 2                   | 8                    |


Analyze the data. Make sure you include

-   null hypothesis (1 point)

-   alternative hypothesis (1 point)

-   explanation for test you will use (1.5 point)

-   results from statistical test (1 point)

-   clear explanation of how results relate to your stated hypotheses (1.5 point)


