---
title: "2024 Fall 78001/78002 Exam Creations"
bibliography: ref_exam_fall_2024.bib
format:
  html:
    embed-resources: true
    toc: true
    toc-float: true
    toc-depth: 6
    code-fold: false
editor:
  visual:
    wrap: 72
  source:
    wrap: 72
---

## Instructions! Read this first.

The exam is open note, open course website, open class-related code
repositories (mine and those you produced in class). However, you may
not get external help ( from other students, directed internet searches,
etc.). Please update the statement below to acknowledge these
instructions (and that you can use git).

I, INSERT YOUR NAME HERE, promise to not seek external help on the exam.
I understand any academic issues will result in a D or F on the exam or
in the class and be reported to the Dean of Students. Infractions will
also result in me being unable to obtain a letter of recommendation from
the department for professional school applications.

For all questions, the statement is based on what you might hear from a colleague 
(e.g., "How do I analyze this?"). This is like when you get a call to review a
paper. Please answer questions succinctly but with necessary details on all
required steps/concerns you might need to consider.

The exam should be uploaded to Blackboard by Friday, 12/20/24, at noon.
If you have issues with Blackboard you can also email it to me, but
please upload to Blackboard if at all possible (this is CUNY policy
unfortunately).

Good luck!

## Background

Southern resident orcas (*Orcinus orca*) in the Pacific Northwest have
recently been spotted wearing salmon on their heads.

[![Photograph By Ocean Wise, DFO Marine Mammal License,
MML18](images/whale_hat.png)](https://www.nationalgeographic.com/animals/article/orcas-puget-sound-salmon-hats-killer-whales)

This behavior was also observed in the 1980's, when it spread among pods
[@whitehead2004], but has rarely been observed since.

Since this is an unusual event, there is interest in both describing it
and trying to determine why it is happening. Long-term data on orcas
make some of this possible, as does community science. A graduate
student friend who is interested in this topic is mining the data. They
ask you for help with analysis.

## Making the data for first few questions

Need to make here and then export for exams and solutions

```{r}
set.seed(16)
whale_yearly_data <-data.frame(year=1987:2024)
whale_yearly_data$population_size <- round(runif(nrow(whale_yearly_data), 60,90))
for (i in 1:nrow(whale_yearly_data)){
  if (whale_yearly_data$year[i] != 1987 & whale_yearly_data$year[i]!=2024){
    whale_yearly_data$hat_wearers[i] <- rbinom(1, whale_yearly_data$population_size[i], .05)
  }
  else(whale_yearly_data$hat_wearers[i] <- rbinom(1, whale_yearly_data$population_size[i], .23))
}
whale_yearly_data$prop_wearing_salmon <- whale_yearly_data$hat_wearers/whale_yearly_data$population_size
whale_yearly_data$salmon_status <- factor(sample(c("high", "low", "closed"),nrow(whale_yearly_data), replace = T))
```

```{r}
saveRDS(whale_yearly_data, "whale_yearly_data.RDS")
remove(whale_yearly_data)
```

```{r}
whale_yearly_data <- readRDS("whale_yearly_data.RDS")
```


## 1

Although "salmon hats" have often been described as fad-based events
that only last for short periods of time, your friend finds this claim
has not been investigated. As noted by Monika Wieland Shields, director
of the Orca Behavior Institute in Washington [@bittel2024],

> “In my opinion it's a stretch to say it was a salmon hat, and an even
> greater stretch to say the fad is back off a single photo....it’s
> quite possible that this has been part of their behavioral repertoire
> that they’ve been doing since time immemorial, and it was just noticed
> by humans in the ‘80s.”

To consider this, they reviewed photos of Northwest orcas from the past
`{r} nrow(whale_yearly_data)` summers and falls (from the craze of 1987 to the
recent resurgence). In each year, they recorded if each orca in the
local pods (the populations are relatively small) was ever recorded with
a salmon hat. They found that outside of 1987 and 2024, orcas were only
spotted wearing head gear
`{r} round(mean(whale_yearly_data[whale_yearly_data$year != 1987 & whale_yearly_data$year!=2024, "prop_wearing_salmon"]), 2)*100`%
of the time. In 2024, `{r} whale_yearly_data[whale_yearly_data==2024, "hat_wearers"]`
out of `{r} whale_yearly_data[whale_yearly_data==2024, "population_size"]` orcas were
spotted wearing hats. How would you recommend they analyze this data?

Analyze the data. Make sure you include

-   null hypothesis (1 point)

-   alternative hypothesis (1 point)

-   explanation for test you will use (1.5 point)

-   results from statistical test (1 point)

```{r}
binom.test(15, 87, .05)#lab
```

-   clear explanation of how results relate to your stated hypotheses
    (1.5 point)

## 2

In 2024, 8 out of 40 males wore salmon hats, while 7 out of 47 females
wore them. Your friend analyzed the data as a t-test (below is some output they 
share with you)

```{r}
t.test(c(7,8), c(40,47))
```

and was excited to get a p-value close to 0.05. They want to know how
you feel about their analysis (How would you analyze this data to determine if there are any differences
in salmon hats between males and females?) and how they should describe 
results that
are close to 0.05. Answer both these questions, and in doing make sure you
define a p-value.



```{r}
chisq.test(matrix(c(8,7,40-8, 47-7), 2,2, byrow = T)) 
```

## 3

One of the hypothesis is that more orcas wear hats when salmon are more
common (maybe because they have more time to play). Although salmon
biomass each year isn't known, the local fishery has historically had
*high*, *low*, and *closed* (meaning no fishing was allowed due to low
numbers) years. Your friend has this data for each year. How should they
analyze it? If it helps, this is the first few rows of their data

```{r}
head(whale_yearly_data)
```

where

-   *year* is year of observation

-   *population_size* is number of resident whales

-   *hat_wearers* is the number of whales that wore hats at least once

-   *prop_wearing_salmon* is the proportion observed wearing hats at least
    once

-   *salmon_status* is the status of the salmon fishery (high, low, or
    closed)

    ```{r}
    glm_model <- glm(cbind(whale_yearly_data$hat_wearers, whale_yearly_data$population_size-whale_yearly_data$hat_wearers)~salmon_status, whale_yearly_data, family = binomial)
    summary(glm_model)
    library(car)
    Anova(glm_model, type = "III")
    library(multcomp)
    summary(glht(glm_model, linfct = mcp(salmon_status = "Tukey")))

    glm_model_quasi <- glm(cbind(whale_yearly_data$hat_wearers, whale_yearly_data$population_size-whale_yearly_data$hat_wearers)~salmon_status, whale_yearly_data, family = "quasibinomial")
    Anova(glm_model_quasi, type = "III")
    summary(glht(glm_model_quasi, linfct = mcp(salmon_status = "Tukey")))
    ```

## 4

Another theory is that larger whales tend to wear hats. Your friend has
weights for the `{r} whale_yearly_data[whale_yearly_data==2024, "population_size"]`
orcas that wore hats in 2024. Their weights are displayed below

```{r}
set.seed(18)
orca_weights <- data.frame(weight = runif(15, 3500, 6000))
saveRDS(orca_weights, "orca_weights.RDS")
remove(orca_weights)
```

```{r, include=F}
orca_weights <- readRDS("orca_weights.RDS")
```


```{r}
library(ggplot2)
ggplot(orca_weights, aes(x=weight))+
  geom_histogram()+
  labs(x="Weight (kg)", y= "Frequency" )
```

The average weight of an orca is 4077 kg. How could they analyze this
data?  Should they be concerned about the distribution of weights?

```{r}
t.test(orca_weights$weight, mu=4077)
```

## 5

Other things besides wearing hats could influence whale size. Your
friend gathers more data on whether whales wear hats, their gender, and
their weight. This is the first few rows of their data

```{r}
set.seed(22)
whale_data_2024 <- data.frame(hat_wearer = factor(c(rep("yes", 15), rep ("no",87-15))),
                              gender = factor(c(rep("male",8), rep("female",7), rep("male", 40-8),rep("female",47-7))),
                              weight = c(rev(sort(orca_weights$weight)),rnorm(40-8, 5366, 200), rnorm(47-7, 3000, 200)))
saveRDS(whale_data_2024, "whale_data_2024.RDS")
remove(whale_data_2024)
```

```{r}
whale_data_2024 <-readRDS("whale_data_2024.RDS")
```

```{r}
head(whale_data_2024)
```

where

-   *hat wearer* denotes if orca wears a hat

-   *gender* denotes male and female

-   *weight* is orca weight in kg

How would you analyze the data?


```{r}
impact <- lm (weight~hat_wearer*gender, whale_data_2024)
plot(impact)
Anova(impact, type="III")

impact_male <- lm (weight~hat_wearer, whale_data_2024[whale_data_2024$gender=="male",])
plot(impact_male)
Anova(impact_male, type="III")

impact_female <- lm (weight~hat_wearer, whale_data_2024[whale_data_2024$gender=="female",])
plot(impact_female)
Anova(impact_female, type="III")
summary(impact_female)
```



## 6

Graph the data from question 5.**Regardless of what you did for #5 (there are multiple ways to answer it) focus on weight as the response variable for the graph.**



## 7 

Orca weight is hard to obtain. For future years your friend would like
to just use length. They give you a new dataset with that includes length. 

```{r} 
set.seed(22)
whale_data_2024_with_length <- whale_data_2024
whale_data_2024_with_length$length <- c(rnorm(8, 6.2, .3), 
                                        rnorm(7, 5.4, .5),
                                        rnorm(40-8, 6.2, .3), 
                                        rnorm(47-7, 5.4, .5))
saveRDS(whale_data_2024_with_length, "whale_data_2024_with_length.RDS")
remove(whale_data_2024_with_length)
```

```{r}
whale_data_2024_with_length <-readRDS("whale_data_2024_with_length.RDS")
```


where

* *length* is orca length in m

How could they determine if this shortcut is appropriate?

```{r}
lwr <- lm(length~weight, whale_data_2024_with_length)
plot(lwr)
Anova(lwr, type="III")
```
