---
title: "2024 Fall 78001 Midterm Solutions"
bibliography: references.bib
format: docx
toc: false
number-sections: false
highlight-style: github
---

## Instructions! Read this first.

The exam is open note, open course website, open class-related code
repositories (mine and any you have produced). However, you may
not get external help ( from other students, directed internet searches,
etc.). Please update the statement below to acknowledge these
instructions.

I, INSERT YOUR NAME HERE, promise to not seek external help on the exam.
I understand any academic issues will result in a D or F on the exam or
in the class and be reported to the Dean of Students. 

The exam should be uploaded to Blackboard by Wednesday, 10/16, at 11:59 PM.  If
you have issues with Blackboard you can also email it to me, but please upload
to Blackboard if at all possible (this is CUNY policy unfortunately).

Good luck!

## Background

A recent paper [@piotto2024] found that micropredation (when animals eat
only a portion of a target prey) by gulls may be important to survival
in southern right whale (*Eubalaena australis*) calves in Argentina.
These predation events cause lesions to develop on whales that may
increase susceptibilty to infection; attempts to prevent them may reduce
maternal energy stores and time spent feeding and resting.

A new graduate student in your program is interested in this topic and
how it may be mediated by food availability, especially in urban
systems. In 2023 over a quarter of the world's North Atlantic right
whale population was observed off Long Island [@riley2024], and while
kelp gulls *Larus dominicanus* are not found in New York, a sister
species (great black-backed gull (*Larus marinus*)) is. Great
black-backed gulls are known to be extremely aggressive predators. They
are also known to forage at dumps [@cavanaugh].

## 1

The student collected data on whales off the coast of Long Island using
photos.\
The photos showed birds were present in the area and included a few
direct acts of attack. Images of whales showed that 44 out of 117 of
whales have lesions! They are excited to write up their first conference
abstract about their work. In a lab presentation noting their follow-up
study, they state that seabirds lead to lesions in almost a third of
whales. Do you have any helpful feedback to offer on this
interpretation? (you may want to look at question 2 before answering so
you don't answer both here - just explain any basic issues here)

## 2

What might be another way to assess how birds impact the occurrence of
lesions? In answering this focus on experimental design - you don't have
to know how to analyze the data (yet).

## 3

Your friend follows up with more information. In general, whales have
lesions 28% of the time for a variety of reasons. Knowing this, how
would you analyze the original data (44 out of 117 whales show lesions
in their photographs)? What hypothesis would you be testing?

You don't have to give the results, just explain and justify the
analysis procedure as if you were writing it up for the methods portion
of a paper. Make sure you state a null hypothesis.


```{r}
binom.test(44,117,.28)#class
```

## 4

Upon analyzing this data, your friend is confused. The outcome states
that the p-value is 0.02, but also provides a probability of success
is 0.37. They know they are supposed to compare something to
.05...Explain what these values are and how to interpret them to your
friend.

## 5

In a follow-up study, your friend gets information on gull attacks on
whales in areas with low (near dumps, which attract birds to land),
medium (in open waters), and high (near fishing vessel routes) gull
density (n=200 per activity type). They then monitor the stress of
whales in these areas (through measurements of glucocorticoid levels
(GC) obtained by collecting whale feces, following work by @lemos2020)
to see if the whales showed any differences in anxiety in these areas.
The data look something like this

```{r, include=F}
#make data and save as rds
set.seed(19)
whale_stress_lecture <- data.frame(gull_activity = c(rep("low", 200), rep("medium", 200),
                                             rep("high",200)), 
                           GC = c(rbeta(200,1,3)*25,
                           rbeta(200,1,3)*50,
rbeta(200,1,3)*100))

whale_stress_lecture$gull_activity <- factor(whale_stress_lecture$gull_activity, c("low", "medium", "high"))

saveRDS(whale_stress_lecture, "whale_stress_lecture.RDS")
remove(whale_stress_lecture)
```

```{r, echo=F, warning=F, message=F}
whale_stress_lecture <- readRDS("whale_stress_lecture.RDS")
library(ggplot2)
library(ggtext)
ggplot(whale_stress_lecture, aes(x=GC, fill=gull_activity))+
  geom_histogram()+
  facet_wrap(~gull_activity, ncol=1)+
  labs(x= "GC concentration ng g<sup>-1</sup>", 
       fill= "Gull activity in area")+
   theme(
    axis.title.x = element_markdown(),
    axis.title.y = element_markdown()
  )
```

How would you recommend analyzing the data? Please describe the best
option and appropriate steps to take given possible results.

```{r}
whale_stress_lm <- lm(GC~gull_activity, whale_stress_lecture)
plot(whale_stress_lm)
library(car)
Anova(whale_stress_lm, type="III")
library(multcomp)
```

## 6

Another study uses low and high activity levels and compares calf
survival in these areas.

|                 |                     |                      |
|-----------------|---------------------|----------------------|
|                 | *Low gull activity* | *High gull activity* |
| *Calf survived* | 31                  | 21                   |
| *Calf died*     | 2                   | 8                    |

How would you analyze this data? Make sure to include a null hypothesis
and explanation of test you chose, why you chose it, and how to
interpret potential results.

```{r}
chisq.test(matrix(c(31,21,2,8), 2, 2, byrow=T))
fisher.test(matrix(c(31,21,2,8), 2, 2, byrow=T))
```

## References
