---
title: "2025 Spring BIO 2100 Final Creation and Answers"
bibliography: ref_exam_spring_2025.bib
format:
  html:
    embed-resources: true
    toc: true
    toc-float: true
    toc-depth: 6
    code-fold: false
editor:
  visual:
    wrap: 72
  source:
    wrap: 72
---

## Instructions! Read this first.

The exam is open note, open course website, open class-related code
repositories (mine and those you produced in class). However, you may
not get external help ( from other students, directed internet searches,
etc.). Please update the statement below to acknowledge these
instructions (and that you can use git).

I, INSERT YOUR NAME HERE, promise to not seek external help on the exam.
I understand any academic issues will result in a D or F on the exam or
in the class and be reported to the Dean of Students. Infractions will
also result in me being unable to obtain a letter of recommendation from
the department for professional school applications.

The exam should be committed and pushed to github by Friday, 5/16/2025,
at 12:30 PM. You can also commit and push along the way, but please mark
your final commit (when you are finished) with the comment **Finished -
ready to grade!.**

For all questions, the statement is based on what you might hear from a
colleague or client if you were a statistical consultant (e.g., "How do
I analyze this?"). Please answer questions succinctly but with necessary
details on all required steps/concerns you might need to consider.

While the exam is based on a realistic scenario (with citations), some
data and scenarios are fictional and contrived for the exam (i.e., don't
quote all this as fact!).

Good luck!

## Background

Earlier this year a company announced they had "resurrected" dire wolves
by gene editing gray wolf embryos [@newsat2025].

[![Screenshot from Science News announcing dire wolf resurrection
(4/11/2025)](Screenshot%202025-05-13%20152253.png)](https://www.science.org/doi/pdf/10.1126/science.ady0869)

Colossal Biosciences, the company that led the dire wolf project, has
asked you, a summer intern, to help analyze some of their data. It turns
out they actually have produced a lot more wolves than they shared with
the public (this is the made up part of the exam!).

## 1

Dire wolves were larger than today's gray wolves. The average male gray
wolf weighs 31 kg. Data below on 26 adult dire wolves the company has
produced is available using

```{r}
male_dire_weight <- read.csv("male_dire_weight.csv", stringsAsFactors = T)
```

Analyze the data to compare weights of dire and gray wolves. Make sure
you include

-   null hypothesis (1 point)

-   alternative hypothesis (1 point)

-   explanation for test you will use (1.5 point)

-   results from statistical test (1 point)

-   clear explanation of how results relate to your stated hypotheses
    (1.5 point)

## 2

A few years later, the company has produced three "lines" of dire wolves
using different gene editing techniques. Is there any evidence the three
lines differ in weight? Data is available using

```{r}
male_dire_weight_groups <- read.csv("male_dire_weight_groups.csv", stringsAsFactors = T)
```

Analyze the data. Make sure you include

-   null hypothesis (1 point)

-   alternative hypothesis (1 point)

-   explanation for test you will use (1.5 point)

-   results from statistical test (1 point)

-   clear explanation of how results relate to your stated hypotheses
    (1.5 point)

## 3

Dire wolves are also supposed to have smaller snouts relative to their
body length than regular wolves. The company provided data from 25 of
their dire wolves and 26 of the gray wolves that were used to source
embryos. Is there evidence the dire wolves are different than gray
wolves in regards to this relationship? Data is available using

```{r}
snout_body_data <- read.csv("snout_body_data.csv", stringsAsFactors = T)
```

Make sure you include

-   null hypothesis (1 point)

-   alternative hypothesis (1 point)

-   explanation for test you will use (1.5 point)

-   results from statistical test (1 point)

-   clear explanation of how results relate to your stated hypotheses.
    Make sure to specify what your results say about the snout length -
    body length relationship(1.5 point)

## 4

Plot the data from question 3

## 5

The company also wants to know if the proportion of dire wolves born
with white fur (a hallmark of dire wolves) is different than that of the
gray wolves that provided the embryos (ignore any genetic issues here!).
18% of the parent gray wolves were white. For the litter in question, 18
out of 25 wolves were white. How would statistically analyze this?

How would you analyze the data. Make sure you include

-   null hypothesis (1 point)

-   alternative hypothesis (1 point)

-   explanation for test you will use (1.5 point)

-   results from statistical test (1 point)

-   clear explanation of how results relate to your stated hypotheses
    (1.5 point)
