---
title: "2025 Spring BIO 2100 Final Creation and Answers"
bibliography: ref_exam_spring_2025.bib
format:
  html:
    embed-resources: true
    toc: true
    toc-float: true
    toc-depth: 6
    code-fold: false
editor:
  visual:
    wrap: 72
  source:
    wrap: 72
---

## Instructions! Read this first.

The exam is open note, open course website, open class-related code
repositories (mine and those you produced in class). However, you may
not get external help ( from other students, directed internet searches,
etc.). Please update the statement below to acknowledge these
instructions (and that you can use git).

I, INSERT YOUR NAME HERE, promise to not seek external help on the exam.
I understand any academic issues will result in a D or F on the exam or
in the class and be reported to the Dean of Students. Infractions will
also result in me being unable to obtain a letter of recommendation from
the department for professional school applications.

The exam should be committed and pushed to github by Friday, 5/16/2025,
at 12:30 PM. You can also commit and push along the way, but please mark
your final commit (when you are finished) with the comment **Finished -
ready to grade!.**

For all questions, the statement is based on what you might hear from a
colleague or client if you were a statistical consultant (e.g., "How do
I analyze this?"). This is like when you get a call to review a paper.
Please answer questions succinctly but with necessary details on all
required steps/concerns you might need to consider.

While the exam is based on a realistic scenario (with citations), some
data and scenarios are fictional and contrived for the exam (i.e., don't
quote all this as fact!).

Good luck!

## Background

Earlier this year a company announced they had "resurrected" dire wolves
by gene editing gray wolf embryos\@newsat2025.

[![Screenshot from Science News announcing dire wolf resurrection
(4/11/2025)](Screenshot%202025-05-13%20152253.png)](https://www.science.org/doi/pdf/10.1126/science.ady0869)

Colossal Biosciences, the company that led the dire wolf project, has
asked you, a summer intern, to help analyze some of their data. It turns
out they actually have produced a lot more wolves than they shared with
the public (this is the made up part of the exam!).

## 1

Dire wolves were larger than today's gray wolves. The average male gray
wolf weighs 31 kg. Data below on 26 adult dire wolves the company has
produced is available using

```{r}
set.seed(29)
male_dire_weight <- data.frame(mass_kg = c(round(rnorm(26, 38, 4),2)))
write.csv(male_dire_weight, "male_dire_weight.csv", row.names = F)
rm(male_dire_weight)
```

```{r}
male_dire_weight <- read.csv("male_dire_weight.csv", stringsAsFactors = T)
```

Analyze the data. Make sure you include

-   null hypothesis (1 point)

    -   H~0~: the average weight of dire wolves is 31 kg

-   alternative hypothesis (1 point)

    -   H~A~: the average weight of dire wolves is not 31 kg

-   explanation for test you will use (1.5 point)

    -   I would use a one-sample t-test because its continuous data from
        a single population. I also don't know the variance but believe
        the means are normally distributed.

-   results from statistical test (1 point)

    ```{r}
    t.test(male_dire_weight$mass_kg, mu=31)
    ```

    -   clear explanation of how results relate to your stated
        hypotheses (1.5 point)

        -   Given a t~25~=8.95 and corresponding p\<.001, I reject the
            null hypothesis. It appears the dire wolves are larger than
            31 kg.

```{r}
t.test(male_dire_weight$mass_kg, mu=31)
```

## 2

A few years later, the company has produced three "lines" of dire wolves
using different gene editing techniques. Is there any evidence the three
lines differ in mean weight? Data is available using

```{r}
set.seed(29)
male_dire_weight_groups <- data.frame(line = c(rep("a", 26), rep("b", 24), rep("c", 27)), mass_kg = c(round(rnorm(77, 38, 4),2)))
write.csv(male_dire_weight_groups, "male_dire_weight_groups.csv", row.names = F)
rm(male_dire_weight_groups)
```

```{r}
male_dire_weight_groups <- read.csv("male_dire_weight_groups.csv", stringsAsFactors = T)
```

Analyze the data. Make sure you include

-   null hypothesis (1 point)

    -   H~0~: there is no difference in mean/average weight among the
        three lines

    -   H~0~: the average weight of the three lines are equal to each
        other/are the same

-   alternative hypothesis (1 point)

    -   H~A~: there is a difference in mean/average weight among the
        three lines

    -   H~A~: the average weight of the three lines are not equal to
        each other/are the same

-   explanation for test you will use (1.5 point)

    -   I will use an anova/lm to consider how means differ among
        groups.

-   results from statistical test (1 point)

    ```{r}
    group_lm <- lm(mass_kg~line, male_dire_weight_groups)
    plot(group_lm)
    ```

    ```{r}
    library(car)
    Anova(group_lm, type="III")
    ```

-   clear explanation of how results relate to your stated hypotheses
    (1.5 point)

    -   Assumptions for the test were checked and met. Given a resulting
        F~2,74~=1.85 and corresponding p = .17, I fail to reject the
        null hypothesis. Line does not appear to impact average wolf
        weight.

## 3

Dire wolves are also supposed to have smaller snouts relative to their
body length than regular wolves. The company provided data from 25 of
their dire wolves and 26 of the gray wolves that were used to source
embryos. Is there evidence the dire wolves are different than gray
wolves? Data is available using

```{r}
set.seed(19)
snout_body_data <- data.frame(wolf = c(rep("gray", 26), rep("dire", 25)), body_length_cm = c(rep(rnorm(51, 60*2.54, 6))))
snout_body_data$snout_length_cm <- NA
snout_body_data[snout_body_data$wolf=="gray", "snout_length_cm"] <- snout_body_data[snout_body_data$wolf=="gray", "body_length_cm"]*(.2+runif(26,0,1)*.01)
snout_body_data[snout_body_data$wolf=="dire", "snout_length_cm"] <- snout_body_data[snout_body_data$wolf=="dire", "body_length_cm"]*(.1+runif(25,0,1)*.01)
write.csv(snout_body_data, "snout_body_data.csv", row.names = F)
rm(snout_body_data)
```

```{r}
snout_body_data <- read.csv("snout_body_data.csv", stringsAsFactors = T)
```

Make sure you include

-   null hypothesis (1 point)

    -   H~0~: snout length is the same among wolf types

    -   H~0~: snout length is not related to body length

    -   H~0~: the relationship between snout length and body length does
        not differ among wolf types

-   alternative hypothesis (1 point)

    -   H~A~: snout length is not the same among wolf types

    -   H~A~:snout length is related to body length

    -   H~A~: the relationship between snout length and body length does
        differ among wolf types

-   explanation for test you will use (1.5 point)

    -   I will use an ANCOVA because I'm considering a numerical
        response variable and how it is explained by both a numerical
        explanatory variable and a categorical explanatory variable and
        potential interactions among them.

-   results from statistical test (1 point)

    ```{r}
    snout_body_data_lm <- lm(snout_length_cm~body_length_cm*wolf, snout_body_data)
    plot(snout_body_data_lm)
    ```

    Asssumptions appear to be met, so I continued with analysis.

    ```{r}
    Anova(snout_body_data_lm, type="III")
    ```

    There is a significant interaction between body and snout length
    based of wolf type (F~1,47~=27.6, p \<.01), so I considered the
    relationship for each wolf type.

    ```{r}
    snout_body_data_lm_gray <- lm(snout_length_cm~body_length_cm, snout_body_data[snout_body_data$wolf == "gray",])
    plot(snout_body_data_lm_gray)
    ```

    ```{r}
    Anova(snout_body_data_lm_gray, type="III")
    summary(snout_body_data_lm_gray)
    ```

    ```{r}
    snout_body_data_lm_dire <- lm(snout_length_cm~body_length_cm, snout_body_data[snout_body_data$wolf == "dire",])
    plot(snout_body_data_lm_dire)
    Anova(snout_body_data_lm_dire, type="III")
    summary(snout_body_data_lm_dire)
    ```

-   clear explanation of how results relate to your stated hypotheses
    (1.5 point)

    -   Given the significant interaction, I considered the snout-body
        length relationship for each wolf type. Although the
        relationship is significant for both types, it appears that the
        snout length is shorter relative to body length for dire wolves
        compared to gray wolves (slopes of .08 vs .2, respectively; all
        p\<.01).

## 4

Plot the data from question 3

```{r}
library(ggplot2)
ggplot(snout_body_data, aes(x=body_length_cm, y=snout_length_cm, color=wolf))+
  geom_point()+
  geom_smooth(method='lm', se=F)+
  labs(title="Snout and body lengths of wolves",
       subtitle = "Gray and dire differ in snout-body length relationship",
       y="Snout length (cm)",
       x = "Body length (cm)",
       color= "Wolf type")+
  theme_bw()
```

```{r}
ggplot(snout_body_data, aes(x=body_length_cm, y=snout_length_cm, color=wolf))+
  geom_point()+
  geom_smooth(method='lm', se=F)+
  labs(title="Snout and body lengths of wolves",
       subtitle = "Gray and dire differ in snout-body length relationship",
       y="Snout length (cm)",
       x = "Body length (cm)",
       color= "Wolf type")+
  theme_bw()+
  facet_wrap(~wolf)
```

## 5

The company also wants to know if the proportion of dire wolves born
with white fur (a hallmark of dire wolves) is different than that of the
gray wolves that provided the embryos (ignore any genetic issues here!).
18% of the parent gray wolves were white. For the litter in question, 18
out of 25 wolves were white. How would statistically analyze this?

How would you analyze the data. Make sure you include

-   null hypothesis (1 point)

    -   H~0~: proportion of white dire wolves is equal to 18%

-   alternative hypothesis (1 point)

    -   H~A~: proportion of white dire wolves is not equal to 18%

-   explanation for test you will use (1.5 point)

    -   We will analyze this data using a binomial test because we are
        comparing a proportion to a set value

-   results from statistical test (1 point)

    ```{r}
    #could use prop.test instead
    binom.test(18, 25, .18)
    ```

-   clear explanation of how results relate to your stated hypotheses
    (1.5 point)

    -   Given a p-value \<.01, I reject the null hypothesis; it appears
        dire wolves are more likely to be white.
