---
title: "2025 Spring BIO 2100 Midterm Creation and Answers"
bibliography: ref_exam_fall_2024.bib
format:
  html:
    embed-resources: true
    toc: true
    toc-float: true
    toc-depth: 6
    code-fold: false
editor:
  visual:
    wrap: 72
  source:
    wrap: 72
---

## Instructions! Read this first.

The exam is open note, open course website, open class-related code
repositories (mine and those you produced in class). However, you may
not get external help ( from other students, directed internet searches,
etc.). Please update the statement below to acknowledge these
instructions (and that you can use git).

I, INSERT YOUR NAME HERE, promise to not seek external help on the exam.
I understand any academic issues will result in a D or F on the exam or
in the class and be reported to the Dean of Students. Infractions will
also result in me being unable to obtain a letter of recommendation from
the department for professional school applications.

The exam should be committed and pushed to github by Friday, 3/28/2025,
at 1:15 PM. You can also commit and push along the way, but please mark
your final commit (when you are finished) with the comment **Finished -
ready to grade!.**

For all questions, the statement is based on what you might hear from a
colleague (e.g., "How do I analyze this?"). This is like when you get a
call to review a paper. Please answer questions succinctly but with
necessary details on all required steps/concerns you might need to
consider.

While the exam is based on a realistic scenario (with citations), some
data and scenarios are fictional and contrived for the exam (i.e., don't
quote all this as fact!).

Good luck!

## Background

Joro spiders (*Trichonephila clavata*) are coming to New York!

[![Image from Nelsen et al
2023](joro_spider_from_nelsen_2023.jpg)](https://onlinelibrary.wiley.com/doi/abs/10.1002/ece3.10728)

Originally from Asia, this spider was first observed in Georgia, USA,
circa 2010. Recent modeling studies have suggested the species could
thrive in New York [@nelsen2023].

[![Species distribution model (SDM) predictions of Trichonephila
clavata's suitable range in North America based on its native Asian
distribution: (a) General Linear Model, (b) General Additive Model, (c)
Maxent, (d) Random Forest, and (e) Averaged prediction of all four
models, calculated as an equally weighted average across all four
primary models, including T. clavata's introduced range as of December
2022. From Nelsen et al.
2023.](ece310728-fig-0005-m.jpg)](https://onlinelibrary.wiley.com/doi/full/10.1002/ece3.10728)

Given the pending (or already happened?) invasion, ecologists are
studying the Joro spider more and more.

## 1

One explanation for the species rapid spread is the specific population
that arrived in the United States differs from other populations of Joro
spiders. For example, they may differ in their use of *ballooning*.
Ballooning occurs when spiders deploy silk threads to catch air currents
and disperse. In Joro spider populations in Japan, 18% of the hatchlings
disperse via ballooning. An inital study of Joro spider hatchlings in
Georgia indicated 22 out of 85 used ballooning. Does this suggest any
difference in the use of ballooning by Joro spiders in Georgia?

Analyze the data. Make sure you include

-   null hypothesis (1 point)

    -   \*the proportion of spiders balloowing is equal to .18\*

-   alternative hypothesis (1 point)

    -   \*the proportion of spiders balloowing is not equal to .18\*

-   explanation for test you will use (1.5 point)

    -   *Since we have 2 outcomes and are comparing to a set proportion,
        I will use a binomial test.*

-   results from statistical test (1 point)

```{r}
 binom.test(22,85,.18)
```

-   clear explanation of how results relate to your stated hypotheses
    (1.5 point)

    -   *Given a p-value of .06, I fail to reject the null hypothesis.*

## 2

While presenting their work at a conference, another scientist shared
they had done a similar but larger study. They found that 110 out of 425
hatchlings from Georgia used ballooning. What does their data indicate
(you can use the same assumption of 18%)?

Analyze the data. Make sure you include

-   null hypothesis (1 point)

    -   \*the proportion of spiders balloowing is equal to .18\*

-   alternative hypothesis (1 point)

    -   \*the proportion of spiders balloowing is not equal to .18\*

-   explanation for test you will use (1.5 point)

    -   *Since we have 2 outcomes and are comparing to a set proportion,
        I will use a binomial test.*

-   results from statistical test (1 point)

```{r}
 binom.test(110, 425, .18)
```

-   clear explanation of how results relate to your stated hypotheses
    (1.5 point)

    -   *Given a p-value of 0.0005, I reject the null hypothesis.*

## 3

A new student is confused when comparing these studies. Both have the
same signal for ballooning.

### a

What is the signal for ballooning rate in both datasets?

```{r}
110/425
22/85
```

*Both samples shows a signal (proportion ballooning) of .258 \#*

### b

However, they differ in p-value. Why? In answering this make sure you
can explain what a p-value is.

*The samples differ in associated p-values due to sample size. P-value
is the probability of getting the signal you observed or something more
extreme if the null hypothesis is true and you sampled the population
multiple times.*

## 4

The larger study also recorded hatchling sex (recorded as male or
female). Data is shown below

|                         | ***Sex*** |        |
|-------------------------|-----------|--------|
| ***Use of ballooning*** | Male      | Female |
| Ballooned               | 54        | 56     |
| Did not balloon         | 150       | 175    |

: Spider hatchling gender and ballooning

How would you analyze the data to determine if sex impacts ballooning?
Make sure you include

-   null hypothesis (1 point)
    -   *ballooning is not continget on sex*
-   alternative hypothesis (1 point)
    -   *ballooning is continget on sex*
-   explanation for test you will use (1.5 point)
    -   *I will use a* $\chi^2$ test because I am comparing proportions
        among groups (table style data)
-   results from statistical test (1 point)

```{r}
chisq.test(matrix(c(54, 56, 150,165), nrow = 2, byrow = T))
```

-   clear explanation of how results relate to your stated hypotheses
    (1.5 point)

    -   *Since the p value is greater than 0.05, we fail to reject the
        null hypothesis*

## 5

Another, related hypothesis is that the invasive spiders have smaller
hatchlings. Weight data was collected from a population in Georgia and
two populations in Asia (Japan and Korea). It can be loaded using.
Weight is in mg.

```{r}
set.seed(29)
spider_mass <- data.frame(population = c(rep("Georgia", 30), rep("Japan", 29), rep("Korea", 30)), mass_mg = c(rnorm(30, .25, .05), rnorm(30, .35, .07), rnorm(29, .32, .04)))
write.csv(spider_mass, "spider_mass.csv", row.names = F)
rm(spider_mass)
```

```{r}
spider_mass <- read.csv("spider_mass.csv", stringsAsFactors = T)
```

How would you analyze the data. Make sure you include

-   null hypothesis (1 point)

    -   *the means of the three populations are equal*

-   alternative hypothesis (1 point)

    -   *the means of the three populations are not equal*

-   explanation for test you will use (1.5 point)

    -   *Given we are comparing means among 3 groups, I will use an
        ANOVA*

-   results from statistical test (1 point)

    -   *First I build the model and test assumptions*

```{r}
spider_mass_model <- lm(mass_mg~population, spider_mass)
plot(spider_mass_model)
```

```         
* *Since assumptions are met, I consider the results*
```

```{r}
library(car)
Anova(spider_mass_model, type="III")
```

```         
* *which indicated population (F~2,86~=16.87, p<.01) has an impact on ballooing rates. Given this, I used post hoc tests to determine which means
differed from others; I used the Tukey method to compare all pairs while controlling for the family-wise error rate.*
```

```{r}
library(multcomp)
spider_mass_compare <- glht(spider_mass_model,  linfct = mcp(population = "Tukey"))
summary(spider_mass_compare)
```

-   clear explanation of how results relate to your stated hypotheses
    (1.5 point)

    -   *the overall p-value ((F~2,86~=16.87, p\<.01) led me to reject
        the null hypothesis. Post hoc analysis indicated all groups
        differed from each other (all p \<.05).* \## 6

Plot the data from question 5.

*Several options here*

```{r}
library(Rmisc)
library(ggplot2)
spider_summary <- summarySE(spider_mass, measurevar="mass_mg", groupvars="population")
ggplot(spider_summary, aes(x=population, y=mass_mg)) +
    geom_point(size=3) +
    geom_errorbar(aes(ymin=mass_mg - ci, ymax=mass_mg + ci), width=0) +
    ylab("Hatchling mass (mg)") +
    xlab("Population") +
    ggtitle("Hatchling Mass by Population") +
    theme(axis.title.x = element_text(face="bold", size=28),
          axis.title.y = element_text(face="bold", size=28),
          axis.text.x = element_text(size=20),
          axis.text.y = element_text(size=20),
          plot.title = element_text(hjust=0.5, face="bold", size=32))
```

```{r}
ggplot(spider_mass, aes(y=mass_mg, x=population, color=population)) +
  geom_jitter() +
  labs(title="Spider Masses in mg from Japan, Korea, and Georgia",
       y= "Spider weight (mg)",
       x= "population")+
  guides(color=F)
```
